---
title: Prop50 impact analysis
format: 
    dashboard: 
        theme : zephyr
        card-style: none
author : Sacha Bondeville
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(boot)
library(sf)
library(leaflet)
library(plotly)
library(shiny)
library(downloadthis)

sv_precinct <- read.csv("/Users/sachabondeville/Documents/stat133/Projects/Project-5/gerrymandering-sachabdvl/data/cleaned_svprecinct.csv")

district_level <-  sv_precinct |>
    group_by(DISTRICT)|>
    summarize(DEM01 = sum(DEM01, na.rm= T),
    DEM02 = sum(DEM02, na.rm = T),
    REP02 = sum(REP02, na.rm = TRUE),
    REP01 = sum(REP01, na.rm = T),
    TOTVOTE = sum(TOTVOTE, na.rm =T))|>
        mutate(WINNER= ifelse(DEM01 > REP01, "DEM", "REP"),
        SCALE = (DEM01+DEM02)/TOTVOTE)
    
# Now we can just compute our score : 

district_vi <- district_level |>
    filter(DEM02 == 0 & REP02 == 0)|>
    mutate(v_i = DEM01 / (TOTVOTE))

mean_median <- district_vi |> 
    summarize(democratscore = mean(v_i)-median(v_i))


n_wasted <- function(by, against) {
    w_threshold <- floor((by + against) / 2) + 1
    n_wasted <- ifelse(by > against, by - w_threshold, by)
    n_wasted
}
ES <- district_vi |>
mutate(wasted_D = n_wasted(DEM01, REP01),
      wasted_R = n_wasted(REP01, DEM01))|>
summarize(EG = (sum(wasted_D)+sum(wasted_R))/(sum(TOTVOTE)))


districts <- st_read("/Users/sachabondeville/Documents/stat133/Projects/Project-5/gerrymandering-sachabdvl/data/Districts/US_Congressional_Districts.shp", quiet = TRUE)

districts <- districts |>
    mutate(DISTRICT = as.double(GEOID))

districts <- st_sf(left_join(districts, district_level, by = join_by(DISTRICT == DISTRICT)))


static_map <- districts |>
    ggplot(aes(fill = WINNER))+geom_sf()+
    scale_fill_manual(
    values = c(
      "DEM" = "#6090F7",
      "REP" = "#FA6E52"))+ 
    theme_minimal(base_family = "Arial")+
    scale_x_continuous(breaks = c())+
    scale_y_continuous(breaks = c())+
    labs(title = "2024 Gen. Election actual results", 
    x = "", 
    y = "", 
    fill = "",
    caption = "Source: Official State data")

districts <- st_transform(districts, crs = 4326)

#pal_continuous <- colorNumeric(
# palette = colorRampPalette(c("#ff0000ff", "#000dffff"))(100),
#  domain = districts$SCALE
#)

pal_discrete <- colorFactor(
palette = c(
  "DEM" = "#6090F7",
  "REP" = "#FA6E52" 
),
domain = districts$WINNER
)

interact_map <- leaflet(districts) |>
addProviderTiles("CartoDB.Positron")|>
addPolygons( 
fillColor = ~pal_discrete(WINNER),
color = "black",
weight = 1,
fillOpacity = 0.8,
label = ~paste0("DEM Votes:", DEM01, " | REP votes : ", REP01))|>
addLegend(
pal = pal_discrete,
values = ~WINNER,
title = "Result",
position = "topright"
)

ab604 <- st_read("/Users/sachabondeville/Documents/stat133/Projects/Project-5/gerrymandering-sachabdvl/data/AB604/AB604.shp", quiet= T)

ab604 <- ab604 |>
    mutate(DISTRICT = as.double(DISTRICT))

sr_precinct <- read.csv("/Users/sachabondeville/Documents/stat133/Projects/Project-5/gerrymandering-sachabdvl/data/cleaned_srprecinct.csv")

block_prop <- sr_precinct |>
    mutate(PROP_DEM_BLOCK = (DEM/sum(DEM))*(PCTSRPREC/100),
    PROP_REP_BLOCK = (REP/(sum(REP)))*(PCTSRPREC/100),
    .by = SRPREC_KEY)|>
    mutate(BLOCK_WEIGHTED_DEM = (DEM01+DEM02)*PROP_DEM_BLOCK,
    BLOCK_WEIGHTED_REP = (REP01+REP02)*PROP_REP_BLOCK)

hypo_districts <- block_prop |>
    group_by(NEWDISTRICT)|>
    summarize(DEM_est = sum(BLOCK_WEIGHTED_DEM, na.rm = T),
    REP_est = sum(BLOCK_WEIGHTED_REP, na.rm = T))|>
    mutate(WINNER= ifelse(DEM_est > REP_est, "DEM", "REP"))|>
    drop_na(NEWDISTRICT)
hypo_district_vi <- hypo_districts |>
    mutate(v_i = DEM_est / (REP_est + DEM_est))

hypo_mean_median <- hypo_district_vi |> 
    summarize(democratscore = mean(v_i)-median(v_i))

hypo_ES <- hypo_district_vi |>
mutate(wasted_D = n_wasted(DEM_est, REP_est),
      wasted_R = n_wasted(REP_est, DEM_est))|>
summarize(EG = (sum(wasted_D)+sum(wasted_R))/(sum(DEM_est + REP_est)))

hypo_districts <- st_sf(left_join(hypo_districts, ab604, by = join_by(NEWDISTRICT == DISTRICT)))

hypo_districts <- st_transform(hypo_districts, crs = 4326)


static_map_hypo <- hypo_districts |>
    ggplot(aes(fill = WINNER))+geom_sf()+
    scale_fill_manual(
    values = c(
      "DEM" = "#6090F7",
      "REP" = "#FA6E52"))+ 
    theme_minimal(base_family = "Arial")+
    scale_x_continuous(breaks = c())+
    scale_y_continuous(breaks = c())+
    labs(title = "2024 Gen. Election hypothetical results", 
    x = "", 
    y = "", 
    fill = "",
    caption = "Source: Official State data")

hypo_districts <- st_transform(hypo_districts, crs = 4326)

#pal_continuous <- colorNumeric(
# palette = colorRampPalette(c("#ff0000ff", "#000dffff"))(100),
#  domain = districts$SCALE
#)

pal_discrete_2 <- colorFactor(
palette = c(
  "DEM" = "#6090F7",
  "REP" = "#FA6E52" 
),
domain = hypo_districts$WINNER
)

interact_map_hypo <- leaflet(hypo_districts) |>
addProviderTiles("CartoDB.Positron")|>
addPolygons( 
fillColor = ~pal_discrete(WINNER),
color = "black",
weight = 1,
fillOpacity = 0.8,
label = ~paste0("DEM Votes:", floor(DEM_est), " | REP votes : ", floor(REP_est)))|>
addLegend(
pal = pal_discrete_2,
values = ~WINNER,
title = "Results",
position = "topright"
)

n_seats_hypo <- sum(hypo_districts$WINNER == "DEM")
n_seats <- sum(district_level$WINNER == "DEM")


```

# Metric results

## Actual 2024 Metric {height = 20%}

```{r}
#| component: valuebox
#| title: "Actual Mean-Median Score"
list(
    color = "success",
    value = mean_median[[1]]
    )
```

```{r}
#| component: valuebox
#| title: Actual number of Democrat seats
list(
  color = "success",
  value = n_seats
)
```

```{r}
#| component: valuebox
#| title: "New hypothetical Mean-Median Score"
list(
    color = "warning",
    value = hypo_mean_median[[1]]
    )
```

```{r}
#| component: valuebox
#| title: Hypothetical number of Democrat seats
list(
  color = "warning",
  value = n_seats_hypo
)

```

## Maps {height = 75%}

```{r}
#| title: "Actual 2024 election results in California"
interact_map
```

```{r}
#| title: "Hypothetical 2024 election results given new districts"
interact_map_hypo
```

##

```{r}
#| title : "Download the map"

ggsave("static_map.png", plot = static_map, width = 6, height = 4, dpi = 300)

ggsave("hypo_static_map.png", plot = static_map_hypo, width = 6, height = 4, dpi = 300)

download_file("static_map.png", button_label = "⬇️ Download plot", button_type = "success")
```

```{r}
download_file("hypo_static_map.png", button_label = "⬇️ Download plot", button_type = "success")
```


# Election Insights


## Introduction {height = 10%} 

Throughout the course of my analysis, I found some an interesting patterns at district and county levels. 

## Association {height = 80%}

```{r}
#| echo: false

insight <- read.csv("/Users/sachabondeville/Documents/stat133/Projects/Project-5/gerrymandering-sachabdvl/data/cleaned_svprecinct.csv")

county_level <-  insight |>
    filter(!(DEM02 > 0),
    !(REP02 > 0))|>
    group_by(COUNTY, DISTRICT)|>
    summarize(DEM = sum(DEM01, na.rm= T), 
    REP = sum(REP01, na.rm = T),
    TOTVOTE = sum(TOTVOTE, na.rm = T))|>
        mutate(WINNER= ifelse(DEM > REP, "DEM", "REP"))

m1 <- lm((REP)~(DEM), data = county_level)

g1 <-  county_level |>
    ggplot(aes(x = DEM, y = REP))+
    geom_smooth(method = "lm", color = "lightblue", se = F)+
    geom_point(aes(text = paste(
    "COUNTY:", COUNTY,
    "<br>Democrat votes:", DEM,
    "<br>Republican votes:", REP,
    "<br>Total votes:", TOTVOTE), color = WINNER))+
    scale_color_manual(
    values = c(
      "DEM" = "#6090F7",
      "REP" = "#FA6E52"))+ 
    theme_minimal(base_family = "Arial")+
    labs(title = "Association: dem and rep vote (county lvl)", 
    x = "# of Democrat votes", 
    y = "# of Republican votes", 
    caption = paste("the slope coefficient of our linear model is of :",round(m1[[1]][2], 4)))

district_level <-  sv_precinct |>
    filter(!(DEM02 > 0),
    !(REP02 > 0))|>
    group_by(DISTRICT)|>
    summarize(DEM = sum(DEM01, na.rm= T), 
    REP = sum(REP01, na.rm = T),
    TOTVOTE = sum(TOTVOTE, na.rm = T))|>
        mutate(WINNER= ifelse(DEM > REP, "DEM", "REP"))

m2 <- lm((REP)~(DEM), data = district_level)



g2 <-  district_level |>
    ggplot(aes(x = DEM, y = REP))+
    geom_smooth(method = "lm", color = "lightblue", se = F)+
    geom_point(aes(text = paste(
    "DISTRICT:", DISTRICT,
    "<br>Democrat votes:", DEM,
    "<br>Republican votes:", REP,
    "<br>Total votes:", TOTVOTE), color = WINNER))+
    scale_color_manual(
    values = c(
      "DEM" = "#6090F7",
      "REP" = "#FA6E52"))+ 
    theme_minimal(base_family = "Arial")+
    labs(title = "Association: dem and rep vote (district lvl)", 
    x = "# of Democrat votes", 
    y = "# of Republican votes", 
    caption = paste("the slope coefficient of our linear model is of :",round(m2[[1]][2], 4)))

```

```{r}
ggplotly(g1)
```

```{r}
ggplotly(g2)
```

## Explanation {height = 10%}

Note the county with the *the highest number of republican votes given its democratic vote* (aka the highest residual) is the **Orange County**. Living in California, a lot of American indeed mentioned that the Orange county was known to be quite republican for LA. I find that is a founded statement indeed! 

## Some other metrics

```{r}
#| component: valuebox
#| title: "Efficiency Score for California 2024 General Election :"
list(
    color = "success",
    value = ES[[1]],4
    )
```

```{r}
#| component: valuebox
#| title: "Hypotetical Efficiency Score, given new district mapping :"

list(
    color = "warning",
    value = hypo_ES[[1]],4
)

```

# Methodology

## Sources

Here are where I found the datasets used for this analysis : 

-   California Statewide Special Election (<https://www.sos.ca.gov/elections/upcoming-elections/statewide-special-nov-4-2025>): official website for the election from the CA Secretary of State.
-   Statewide Database Precinct Data (<https://statewidedatabase.org/d20/g24.html>): source of precinct-level data from the 2024 General Election.
-   2024 Election Geographic Data (<https://statewidedatabase.org/d20/g24_geo_conv.html>): source of shapefiles for precinct maps from 2024 General Election.
-   Proposed Congressional Map (<https://aelc.assembly.ca.gov/proposed-congressional-map>): source of shapefiles for district map proposed in AB 604.


**Gerrymandering metrics documentation:**

*Mean-Median Score:*
Here is the code used to compute this metric : 

```r

district_vi <- district_level |>
    filter(DEM02 == 0 & REP02 == 0)|>
    mutate(v_i = DEM01 / (TOTVOTE))

mean_median <- district_vi |> 
    summarize(democratscore = mean(v_i)-median(v_i))

```

It follows the following formula : 
Suppose we observe an election where the vote share for Republicans in each of 𝑛 districts is 𝑣1,…,𝑣𝑁.
The mean-median score is:
𝑀=mean(𝑣𝑖)−median(𝑣𝑖)

*Efficiency score:*
Here is the code used to compute this metric : 

```r

ES <- district_vi |>
mutate(wasted_D = n_wasted(DEM01, REP01),
      wasted_R = n_wasted(REP01, DEM01))|>
summarize(EG = (sum(wasted_D)+sum(wasted_R))/(sum(TOTVOTE)))

```
It follows the following logic: 

Suppose two parties, A and B. The efficiency gap (EG) favoring party A for a given pattern of votes, is defined as the difference in wasted votes divided by the total number of votes:

$EG = \frac{W_B - W_A}{T}$

Where 𝑊𝐵 is the number of wasted votes by Party B, 𝑊𝐴 is the number of wasted votes by Party A, and 𝑇 is the total votes cast.

## Data Cleaning methodology

Here is the code used for the cleaning process. It mostly just relies on simple pattern recognition, and column selection:

<button onclick="window.open('https://sacha-bondeville.quarto.pub/data-cleaning-gerrymandering/', '_blank')" 
        style="padding:10px 20px; background-color:#0078d7; color:white; border:none; border-radius:6px; cursor:pointer;">
  📄 Open PDF Report
</button>