---
title: Exploratory Data Analysis
format: html
author : Sacha Bondeville
---

```{r}
library(tidyverse)
library(boot)
library(plotly)
sv_precinct <- read.csv("Projects/Project-5/gerrymandering-sachabdvl/data/cleaned_svprecinct.csv")

```


### Question 1

*What is the total distribution of vote counts at **the county level** ?*


### Answer 1

```{r}

county_level <-  sv_precinct |>
    group_by(COUNTY)|>
    summarize(DEM01 = sum(DEM01, na.rm= T), 
    REP01 = sum(REP01, na.rm = T),
    TOTVOTE = sum(TOTVOTE))|>
        mutate(WINNER= ifelse(DEM01 > REP01, "DEM", "REP"))

county_level |>
    ggplot(aes(x = TOTVOTE))+geom_histogram(fill = "goldenrod")+ 
    labs(title = "Distribution of votes", 
    x = "#Total votes", 
    y  = "Count")+
    theme_minimal(base_family = "Arial")

```


### Question 2 : 
What is the association between number of democrat vote and number of republican vote at the county level ? 

### Answer 2: 

```{r}



m1 <- lm(REP01~DEM01, data = county_level)

g1 <-  county_level |>
    ggplot(aes(x = DEM01, y = REP01))+
    geom_smooth(method = "lm", color = "lightblue", se = F)+
    geom_point(aes(text = paste(
    "COUNTY:", COUNTY,
    "<br>Democrat votes:", DEM01,
    "<br>Republican votes:", REP01,
    "<br>Total votes:", TOTVOTE), color = WINNER))+
    scale_color_manual(
    values = c(
      "DEM" = "#6090F7",
      "REP" = "#FA6E52"))+ 
    theme_minimal(base_family = "Arial")+
    labs(title = "Association: democrat and republican vote", 
    x = "# of Democrat votes", 
    y = "# of Republican votes", 
    caption = paste("the slope coefficient of our linear model is of :",round(m1[[1]][2], 4)))

ggplotly(g1)
```

### Question 3: 

*What is the county with the highest number of republican votes given their number of democrat votes ?* 

### Answer 3: 

```{r}
county_level |>
    mutate(residual = resid(m1))|>
    arrange(desc(residual))|>
    select(COUNTY)|>
    slice(1)
```

The Orange County has the highest number of republican votes given democrat votes, which is not that surprising given the disparity of its social groups. 







