---
title: Gerrymandering Metrics
format: typst
---

```{r}
#| echo: false 
#| warning: false
library(tidyverse)
library(boot)
sv_precinct <- read.csv("Projects/Project-5/gerrymandering-sachabdvl/data/cleaned_svprecinct.csv")

```

# 2024 Election Results and the 2024 District Map

## Metrics

### Mean-Median approach: 

```{r}
district_level <-  sv_precinct |>
    group_by(DISTRICT)|>
    summarize(DEM01 = sum(DEM01, na.rm= T), 
    REP01 = sum(REP01, na.rm = T),
    TOTVOTE = sum(TOTVOTE))

# Now we can just compute our score : 

district_vi <- district_level |>
    mutate(v_i = DEM01 / (TOTVOTE))

district_vi |> 
    summarize(democratscore = mean(v_i)-median(v_i))
```

The mean-median score is of -0.00759. But we don't really know what to compare it to! It seems there is a very small negative score in defavor of democrats. 

#### Side quest : *Is the mean median score statistically significant ?*

```{r}
# We define a function that can compute that systematically on a bootstrap test: 

democratscore_fn <- function(data, index) {
  d <- data[index, ]
  mean(d$v_i) - median(d$v_i)
}
set.seed(123)
boot_res <- boot(district_vi, democratscore_fn, R = 2000)

boot.ci(boot_res, type = "perc")
```

The CI interval reveals there is no statistical significance on this measure ! 

### Efficiency Score

```{r}
n_wasted <- function(by, against) {
    w_threshold <- floor((by + against) / 2) + 1
    n_wasted <- ifelse(by > against, by - w_threshold, by)
    n_wasted
}
district_level |>
mutate(wasted_D = n_wasted(DEM01, REP01),
      wasted_R = n_wasted(REP01, DEM01))|>
summarize(EG = (sum(wasted_D)+sum(wasted_R))/(sum(TOTVOTE)))

```


## Map ! 





# 2024 Election Results and the proposed 2025 District Map

