---
title: Gerrymandering Metrics
format: typst
---

```{r}
#| echo: false 
#| warning: false
library(tidyverse)
library(boot)
library(sf)
library(leaflet)
library(plotly)

sr_precinct <- read.csv("/Users/sachabondeville/Documents/stat133/Projects/Project-5/gerrymandering-sachabdvl/data/cleaned_srprecinct.csv")
```

# 2024 Election Results and the 2024 District Map

## Metrics

### Mean-Median approach: 
Remember we need to disregard districts in which there was a choice between two democrats or two republicans ! 
```{r}
district_level <-  sr_precinct |>
    group_by(DISTRICT)|>
    summarize(DEM01 = sum(DEM01, na.rm= T),
    DEM02 = sum(DEM02, na.rm = T),
    REP02 = sum(REP02, na.rm = TRUE),
    REP01 = sum(REP01, na.rm = T),
    TOTVOTE = sum(TOTVOTE, na.rm =T))|>
        mutate(WINNER= ifelse(DEM01 > REP01, "DEM", "REP"),
        SCALE = (DEM01+DEM02)/TOTVOTE)
    
# Now we can just compute our score : 

district_vi <- district_level |>
    filter(DEM02 == 0 & REP02 == 0)|>
    mutate(v_i = DEM01 / (TOTVOTE))|>
    drop_na(v_i)

mean_median <- district_vi |> 
    summarize(democratscore = mean(v_i)-median(v_i))

mean_median
```

The mean-median score is of -0.00125. But we don't really know what to compare it to! It seems there is a very small negative score in defavor of democrats. 

#### Side quest : *Is the mean median score statistically significant ?*

```{r}
# We define a function that can compute that systematically on a bootstrap test: 

democratscore_fn <- function(data, index) {
  d <- data[index, ]
  mean(d$v_i) - median(d$v_i)
}
set.seed(123)
boot_res <- boot(district_vi, democratscore_fn, R = 2000, )

boot.ci(boot_res, type = "perc")
```

The CI interval reveals there is no statistical significance on this measure ! 

### Efficiency Score

```{r}
n_wasted <- function(by, against) {
    w_threshold <- floor((by + against) / 2) + 1
    n_wasted <- ifelse(by > against, by - w_threshold, by)
    n_wasted
}
ES <- district_level |>
mutate(wasted_D = n_wasted(DEM01, REP01),
      wasted_R = n_wasted(REP01, DEM01))|>
summarize(EG = (sum(wasted_D)+sum(wasted_R))/(sum(TOTVOTE)))

ES

```


## Map ! 

```{r}
districts <- st_read("/Users/sachabondeville/Documents/stat133/Projects/Project-5/gerrymandering-sachabdvl/data/Districts/US_Congressional_Districts.shp")

districts <- districts |>
    mutate(DISTRICT = as.double(GEOID))

districts <- st_sf(left_join(districts, district_level, by = join_by(DISTRICT == DISTRICT)))

```

```{r}

districts |>
    ggplot(aes(fill = WINNER))+geom_sf()+
    scale_fill_manual(
    values = c(
      "DEM" = "#6090F7",
      "REP" = "#FA6E52"))+ 
    theme_minimal(base_family = "Arial")+
    scale_x_continuous(breaks = c())+
    scale_y_continuous(breaks = c())+
    labs(title = "2024 Gen. Election actual results", 
    x = "", 
    y = "", 
    fill = "",
    caption = "Source: Official State data")

districts <- st_transform(districts, crs = 4326)

#pal_continuous <- colorNumeric(
# palette = colorRampPalette(c("#ff0000ff", "#000dffff"))(100),
#  domain = districts$SCALE
#)

pal_discrete <- colorFactor(
palette = c(
  "DEM" = "#6090F7",
  "REP" = "#FA6E52" 
),
domain = districts$WINNER
)



```


# 2024 Election Results and the proposed 2025 District Map

## Methodology : 

```{r}
#| echo: false
ab604 <- st_read("/Users/sachabondeville/Documents/stat133/Projects/Project-5/gerrymandering-sachabdvl/data/AB604/AB604.shp")

ab604 <- ab604 |>
    mutate(DISTRICT = as.double(DISTRICT))



```

We use the registration vote to estimate the amount of DEM votes and REP votes within each block. This estimation will then allow us to study the impact of PROP50 redistribution ! 
```{r}
block_prop <- sr_precinct |>
    mutate(PROP_DEM_BLOCK = (DEM/sum(DEM))*(PCTSRPREC/100),
    PROP_REP_BLOCK = (REP/(sum(REP)))*(PCTSRPREC/100),
    .by = SRPREC_KEY)|>
    mutate(BLOCK_WEIGHTED_DEM = (DEM01+DEM02)*PROP_DEM_BLOCK,
    BLOCK_WEIGHTED_REP = (REP01+REP02)*PROP_REP_BLOCK)
```

Now we have to redistribute these blocks with the new district mapping.

```{r}
hypo_districts <- block_prop |>
    group_by(NEWDISTRICT)|>
    summarize(DEM_est = sum(BLOCK_WEIGHTED_DEM, na.rm = T),
    REP_est = sum(BLOCK_WEIGHTED_REP, na.rm = T))|>
    mutate(WINNER= ifelse(DEM_est > REP_est, "DEM", "REP"))|>
    drop_na(NEWDISTRICT)
```

## Metrics : 

### Mean-Median approach: 
*Following the same methodology as above :*

```{r}
hypo_district_vi <- hypo_districts |>
    mutate(v_i = DEM_est / (REP_est + DEM_est))

hypo_mean_median <- hypo_district_vi |> 
    summarize(democratscore = mean(v_i)-median(v_i))

hypo_mean_median
```

The mean-median score increased slightly with Prop50

### Side quest : *Is the mean median score statistically significant ?*

```{r}
set.seed(43)
boot_res_2 <- boot(hypo_district_vi, democratscore_fn, R = 2000)

boot.ci(boot_res_2, type = "perc")
```

The mean-median score increased in favor of the democrats, but the result is still not exactly statistically significant. This reveals the weakness of that measure, as PROP50 *intentionally* redistricted, which should have affected the metric more. 

### Efficiency Score : 

```{r}
hypo_ES <- hypo_districts |>
mutate(wasted_D = n_wasted(DEM_est, REP_est),
      wasted_R = n_wasted(REP_est, DEM_est))|>
summarize(EG = (sum(wasted_D)+sum(wasted_R))/(sum(DEM_est + REP_est)))

hypo_ES
```

The efficiency score nearly did not change at all ! 

## Hypothetical map : 

```{r}
hypo_districts <- st_sf(left_join(hypo_districts, ab604, by = join_by(NEWDISTRICT == DISTRICT)))

hypo_districts <- st_transform(hypo_districts, crs = 4326)


hypo_districts |>
    ggplot(aes(fill = WINNER))+geom_sf()+
    scale_fill_manual(
    values = c(
      "DEM" = "#6090F7",
      "REP" = "#FA6E52"))+ 
    theme_minimal(base_family = "Arial")+
    scale_x_continuous(breaks = c())+
    scale_y_continuous(breaks = c())+
    labs(title = "2024 Gen. Election hypothetical results", 
    x = "", 
    y = "", 
    fill = "",
    caption = "Source: Official State data")

hypo_districts <- st_transform(hypo_districts, crs = 4326)

#pal_continuous <- colorNumeric(
# palette = colorRampPalette(c("#ff0000ff", "#000dffff"))(100),
#  domain = districts$SCALE
#)

pal_discrete_2 <- colorFactor(
palette = c(
  "DEM" = "#6090F7",
  "REP" = "#FA6E52" 
),
domain = hypo_districts$WINNER
)
```

# Interactive maps code : 

```{r}
#| eval: false

# Actual results: 
leaflet(districts) |>
addProviderTiles("CartoDB.Positron")|>
addPolygons( 
fillColor = ~pal_discrete(WINNER),
color = "black",
weight = 1,
fillOpacity = 0.8,
label = ~paste0("DEM Votes:", DEM01, " | REP votes : ", REP01))|>
addLegend(
pal = pal_discrete,
values = ~WINNER,
title = "Election result",
position = "bottomright"
)

#Hypothetical results: 
leaflet(hypo_districts) |>
addProviderTiles("CartoDB.Positron")|>
addPolygons( 
fillColor = ~pal_discrete(WINNER),
color = "black",
weight = 1,
fillOpacity = 0.8,
label = ~paste0("DEM Votes:", floor(DEM_est), " | REP votes : ", floor(REP_est)))|>
addLegend(
pal = pal_discrete_2,
values = ~WINNER,
title = "Election result",
position = "bottomright"
)

```

