---
title: "Practice Worksheet: Gerrymandering"
format: typst
---

Recall the 3 quantitative approaches to measuring gerrymandering discussed in lecture:\
1. Mean - Median Score (Grofman, 1983)\
2. Efficiency Gap (Stephanopolous and McGee, 2015)\
3. Ensembles / Extreme Outliers

\
Use the following table to practice applying the first two techniques.\

| District  | D Votes | R Votes | Result |
|:---------:|:-------:|:-------:|:-------|
|     1     |   62    |   38    | D wins |
|     2     |   55    |   45    | D wins |
|     3     |   44    |   56    | R wins |
|     4     |   47    |   53    | R wins |
|     5     |   46    |   54    | R wins |
| **Total** | **254** | **246** |        |

```{r}
#| message: false
#| echo: false
library(tidyverse)
votes <- tibble(District = 1:5,
                d_votes = c(62, 55, 44, 47, 46),
                r_votes = c(38, 45, 56, 53, 54),
                result = c(rep("D wins", 2), rep("R wins", 3)))
```

### Q1. Mean-Median Score

Calculate the mean-median score associated with this particular district breakdown of votes.

```{r}

votes |>
mutate(v_i = d_votes / (d_votes + r_votes))|>
summarize(democratscore = mean(v_i)-median(v_i))
```

### Q2. Efficiency Gap

Calculate the Efficiency Gap (EG) associated with this particular district breakdown of votes.

*Tip*: it may be helpful to write a function that takes a vector of votes across districts for party A and vector of votes for the party B and returns a vector of the number votes wasted by party A across districts.

```{r}
n_wasted <- function(by, against) {
    w_threshold <- floor((by + against) / 2) + 1
    n_wasted <- ifelse(by > against, by - w_threshold, by)
    n_wasted
}
votes |>
mutate(wasted_D = n_wasted(d_votes, r_votes),
      wasted_R = n_wasted(r_votes, d_votes))|>
summarize(EG = (sum(wasted_D)+sum(wasted_R))/(sum(d_votes)+sum(r_votes)))



```